High-Dimensional Regularized Discriminant Analysis

### 2013-12-29 reports/paper-hdrda.Rnw
Umm, let's minimize cv error rather than maximize it.
Initial introduction of model-selection algorithm. Also, added a shell of things to close out #11

To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denote the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Let $\bm A \oplus \bm B$
denote the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A) = \{\bm z \in \mathbb{R}_{p \times 1} : \bm A
\bm z = \bm 0_{m \times 1}\}$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Denote by $\mathbb{R}_{p \times p}^{>}$ the cone of real $p \times
p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times p}^{\ge}$
denote the cone of real $p \times p$ positive-semidefinite matrices. Let
$V^{\perp} = \{\bm y \in \mathbb{R}_{p \times 1} : \bm y^{T} \bm v = 0$ for all
$\bm v \in V \}$ denote the orthogonal complement of a vector space $V \subset
\mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if $c \ne 0$
and $0$ otherwise.

@@

\subsection{Properties of the HDRDA Classifier}
\label{sec:computational-efficiency}

@@

\subsection{Computationally Efficient Model Selection}

Conveniently, we see that the inverse and determinant of $\bm W_k$ are each in
terms of the inverse and determinant of $\bm Q_k \in \mathbb{R}_{n_k \times
  n_k}$, respectively, as well as the diagonal matrix $\bm \Gamma_k \in
\mathbb{R}_{q \times q}$. Hence, the resulting formulation requires $O(n_k p q)$
calculations to compute both $\bm W_k^{-1}$ and $| \bm W_k |$ and is especially
advantageous when the number of classes $K$ is large because $n_k$ is then
potentially much smaller than $q$. The expressions given in Proposition
\ref{proposition:hdrda-W_k} also expedite the selection of $\lambda$ and
$\gamma$ via cross-validation because the most time-consuming matrix operation
involved in computing $\bm W_k^{-1}$ and $|\bm W_k|$ is $\bm X_k \bm U_1 \in
\mathbb{R}_{n_k \times q}$, which is independent of $\lambda$ and
$\gamma$. Therefore, we need calculate $\bm X_k \bm U_1$ only once for each
cross-validation fold. The subsequent operations in calculating $\bm W_k^{-1}$
and $|\bm W_K|$ can be simply updated for different pairs of $\lambda$ and
$\gamma$.

Thus far, we have constructed each $\tilde{\bm \Sigma}_k$ in terms of the first
$q$ eigenvectors of $\widehat{\bm \Sigma}$. By utilizing a technique similar to
the so-called kernel trick, we can obtain $\bm U_1$ by computing the
eigendecomposition of a much smaller $N \times N$ matrix when $p \gg N$
\citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix is
$\widehat{\bm \Sigma} = \bm X^{T} \bm X$\footnote{TODO: $1/N$ is missing. Add
  this. How much do things change below?} Then, because $\bm X = \bm M \bm
D^{1/2} \bm U^{T}$ by the singular value decomposition, we have the
eigendecomposition $\widehat{\bm \Sigma} = \bm U \bm D \bm U^{T}$, where $\bm M
\in \mathbb{R}_{N \times N}$ is orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N
  \times N}^{\ge}$ is a diagonal matrix consisting of the singular values of
$\bm X$, and $\bm U \in \mathbb{R}_{p \times N}$ is orthogonal. Now, $\bm X \bm
X^{T} = \bm M \bm D \bm M^{T}$ is the eigenvalue decomposition of $\bm X \bm
X^{T}$, where $\bm D$ is the matrix of eigenvalues of $\bm X \bm X^{T}$ and the
columns of $\bm M$ are the corresponding eigenvectors. Hence, we obtain $\bm M$
and $\bm D$ from an eigendecomposition of $\bm X \bm X^{T} \in \mathbb{R}_{N
  \times N}$. Next, we compute $\bm U = \bm X^{T} \bm M \bm D^{+/2}$, where $\bm
D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We then determine $q$, the number
of numerically nonzero eigenvalues present in $\bm D$, by calculating the number
of eigenvalues that exceeds some tolerance value, say, $\epsilon = 1 \times
10^{-6}$. After calculating $q$, we extract $\bm U_1$ by retaining only the
first $q$ columns of $\bm U$.

\IncMargin{1em}
\begin{algorithm}
  \SetKwData{Left}{left}
  \SetKwData{This}{this}
  \SetKwData{Up}{up}
  \SetKwFunction{Union}{Union}
  \SetKwFunction{FindCompress}{FindCompress}
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}

  \Input{
    Data matrix $\bm X$\\
    Parameter grid $\{\lambda_g\}_{g=1}^G \times \{\gamma_h\}_{h=1}^H$
  }
  \Output{Optimal Estimates, $\hat{\lambda}$ and $\hat{\gamma}$}
  \BlankLine
  \For{$v \leftarrow 1$ \KwTo $V$ \tcp*[r]{$V$-fold cross-validation}} {
    Partition $v$th fold into $\bm X_{train} \in \mathbb{R}_{N \times p}$ and
    $\bm X_{test} \in \mathbb{R}_{N_T \times p}$

@@

      Center $\bm X_k \leftarrow \bm X_k - \bm j_{n_k} \xbar_k^T$    
    }

@@

  Report optimal $(\hat{\lambda}, \hat{\gamma}) \leftarrow \argmax_{(\lambda, \gamma)} \widehat{\text{Error}}(\lambda, \gamma)$
  \caption{Model selection for the \emph{HDRDA} classifier}
  \label{algo_hdrda_model_selection}
\end{algorithm}
\DecMargin{1em}

The compact SVD is sometimes called the reduced or fast SVD.

Let $\bm j_m \in \mathbb{R}_{m \times 1}$ denote a vector of ones.

>>>>

To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denote the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Define $\bm 1_m \in
\mathbb{R}_{m \times 1}$ as a vector of ones. Let $\bm A \oplus \bm B$ denote
the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A) = \{\bm z \in \mathbb{R}_{p \times 1} : \bm A
\bm z = \bm 0_{m \times 1}\}$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Denote by $\mathbb{R}_{p \times p}^{>}$ the cone of real $p \times
p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times p}^{\ge}$
denote the cone of real $p \times p$ positive-semidefinite matrices. Let
$V^{\perp} = \{\bm y \in \mathbb{R}_{p \times 1} : \bm y^{T} \bm v = 0$ for all
$\bm v \in V \}$ denote the orthogonal complement of a vector space $V \subset
\mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if $c \ne 0$
and $0$ otherwise.

@@

\subsection{Properties of the HDRDA Classifier}
\label{sec:hdrda-properties}

@@

\subsection{Model Selection}

Thus far, we have presented the \emph{HDRDA} classifier and its properties that
facilitate an efficient calculation of the decision rule. Here, we describe in
an efficient model-selection procedure to select the optimal estimates for the
tuning parameters $\lambda$ and $\gamma$. We provide pseudocode for the model
selection in Algorithm \ref{alg:hdrda_model_selection}. Our goal is to select
the optimal tuning parameter estimates from the Cartesian product of candidate
values, $\{\lambda_g\}_{g=1}^G \times \{\gamma_h\}_{h=1}^H$. We utilize $V$-fold
cross-validation to estimate the conditional error rate for each candidate
tuning-parameter pair and select the pair that attains the minimum error
rate. To calculate the $V$-fold cross-validation, we partition the original
training data into $V$ mutually exclusive and exhaustive folds that have
approximately the same number of observations. Then, for $v = 1, \ldots, V$, we
classify the observations in the $v$th fold by training a classifier on the
remaining $V - 1$ folds for each candidate pair of tuning parameters. We then
calculate the proportion of misclassified observations across the $V$ folds and
select the pair $(\widehat{\lambda}, \widehat{\gamma})$ that minimizes the
considered training error-rate estimator.

\IncMargin{1em}
\begin{algorithm}
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}

  \Input{
    Data matrix $\bm X$\\
    Parameter grid $\{\lambda_g\}_{g=1}^G \times \{\gamma_h\}_{h=1}^H$
  }
  \Output{Optimal Estimates $(\hat{\lambda}, \hat{\gamma})$}
  \BlankLine
  \For{$v \leftarrow 1$ \KwTo $V$} {
    Partition $\bm X$ into $\bm X_{train} \in \mathbb{R}_{N \times p}$ and
    $\bm X_{test} \in \mathbb{R}_{N_T \times p}$

@@

      Center $\bm X_k \leftarrow \bm X_k - \bm 1_{n_k} \xbar_k^T$    
    }

@@

  Report optimal $(\hat{\lambda}, \hat{\gamma}) \leftarrow \argmin_{(\lambda, \gamma)} \widehat{\text{Error}}(\lambda, \gamma)$
  \caption{Model selection for the \emph{HDRDA} classifier}
  \label{alg:hdrda_model_selection}
\end{algorithm}
\DecMargin{1em}

% TODO: Now that the algorithm has been introduced and described, we need to
% discuss the details. Let's remove some of the bloat where we talk about the
% kernel trick, etc. Simply discuss how this technique is fast and cite Hastie's
% book.

A primary contributing factor to the efficiency of Algorithm
\ref{alg:hdrda_model_selection} is our usage of the compact singular value
decomposition (SVD) (sometimes called the reduced or fast SVD), which we
describe here. Rather than computing the eigenvalue decomposition of
$\widehat{\bm \Sigma}$ to obtain $\bm U_1$, we instead obtain $\bm U_1$ by
computing the eigendecomposition of a much smaller $N \times N$ matrix when $p
\gg N$ \citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix is
$\widehat{\bm \Sigma} = \bm X^{T} \bm X$\footnote{TODO: $1/N$ is missing. Add
  this. It will be absorbed into the eigenvalues.}  Then, because $\bm X = \bm M
\bm D^{1/2} \bm U^{T}$ by the SVD, we have the eigendecomposition $\widehat{\bm
  \Sigma} = \bm U \bm D \bm U^{T}$, where $\bm M \in \mathbb{R}_{N \times N}$ is
orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N \times N}^{\ge}$ is a diagonal matrix
consisting of the singular values of $\bm X$, and $\bm U \in \mathbb{R}_{p
  \times N}$ is orthogonal. Now, $\bm X \bm X^{T} = \bm M \bm D \bm M^{T}$ is
the eigenvalue decomposition of $\bm X \bm X^{T}$, where $\bm D$ is the matrix
of eigenvalues of $\bm X \bm X^{T}$ and the columns of $\bm M$ are the
corresponding eigenvectors. Hence, we obtain $\bm M$ and $\bm D$ from an
eigendecomposition of $\bm X \bm X^{T} \in \mathbb{R}_{N \times N}$. Next, we
compute $\bm U = \bm X^{T} \bm M \bm D^{+/2}$, where $\bm D^{+/2} = \bm
D_q^{-1/2} \oplus \bm 0_{N-q}$. We then determine $q$, the number of numerically
nonzero eigenvalues present in $\bm D$, by calculating the number of eigenvalues
that exceeds some tolerance value, say, $\epsilon = 1 \times 10^{-6}$. After
calculating $q$, we extract $\bm U_1$ by retaining only the first $q$ columns of
$\bm U$.

% TODO: Discuss things that need only be computed once
% TODO: Discuss operations independent of tuning parameters

\subsection{Computational Complexity}

Conveniently, we see that the inverse and determinant of $\bm W_k$ are each in
terms of the inverse and determinant of $\bm Q_k \in \mathbb{R}_{n_k \times
  n_k}$, respectively, as well as the diagonal matrix $\bm \Gamma_k \in
\mathbb{R}_{q \times q}$. Hence, the resulting formulation requires $O(n_k p q)$
calculations to compute both $\bm W_k^{-1}$ and $| \bm W_k |$. The expressions
given in Proposition \ref{proposition:hdrda-W_k} also expedite the selection of
$\lambda$ and $\gamma$ via cross-validation because the most time-consuming
matrix operation involved in computing $\bm W_k^{-1}$ and $|\bm W_k|$ is $\bm
X_k \bm U_1 \in \mathbb{R}_{n_k \times q}$, which is independent of $\lambda$
and $\gamma$. Therefore, we need calculate $\bm X_k \bm U_1$ only once for each
cross-validation fold. The subsequent operations in calculating $\bm W_k^{-1}$
and $|\bm W_K|$ can be simply updated for different pairs of $\lambda$ and
$\gamma$.

\begin{enumerate}
  \item $\bm W_k^{-1}$\footnote{TODO: Need complexity for the following}
  \item $|\bm W_k|$
  \item (12)
\end{enumerate}
